---
title: "Reproducible research - peer graded W2 assignment"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# 1. Reading in the dataset and preparing the data
```{r}
library("data.table")
```

```{r}
zip_url <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"
temp <- tempfile()
download.file(zip_url,temp)
activity_ds <- read.csv(unz(temp, "activity.csv"))
activity_ds <- setDT(activity_ds)
```
# 2. Histogram of the total number of steps taken each day
## 2.1 Calculating the total number of steps taken per day
```{r}
total_num_of_steps <- activity_ds[,.(sum_of_steps = sum(steps)),date]
total_num_of_steps
```

## 2.2 Building histogram of the total number of steps taken each day
```{r}
hist(total_num_of_steps$sum_of_steps, 
     main = "Distribution of total number of steps per day",
     xlab = "Total number of steps")
```

# 3. Calculating the mean and median of the total number of steps taken per day
```{r}
mean_and_median_num_of_steps <- activity_ds[,.(mean_of_steps = mean(steps), median_of_steps = median(steps)),date]
mean_and_median_num_of_steps

mean_steps_across_all_days <- mean(total_num_of_steps$sum_of_steps,na.rm = TRUE)
mean_steps_across_all_days # 10766.19

median_steps_across_all_days <- median(total_num_of_steps$sum_of_steps,na.rm = TRUE)
median_steps_across_all_days # 10765
```

# 4. Time series plot of the average number of steps taken
```{r}
mean_of_steps_across_intervals <- 
  activity_ds[,.(mean_of_steps = mean(steps,na.rm = TRUE)), interval]

plot(mean_of_steps_across_intervals$interval,
     mean_of_steps_across_intervals$mean_of_steps, type = "l",
     main = "Average number of steps taken",
     xlab = "Interval", ylab = "Mean of steps")
```

# 5. Determining the 5-minute interval that, on average, contains the maximum number of steps
```{r}
mean_of_steps_across_intervals[which.max(mean_of_steps_across_intervals$mean_of_steps)]
```

# 6. Code to describe and show a strategy for imputing missing data
## 6.1 Number of missing values in the dataset
```{r}
sum(is.na(activity_ds$steps))
```

## 6.2 Imputting missing values by 5 min interval mean values
```{r}
merged_activity_ds <- merge(activity_ds, mean_of_steps_across_intervals,
                            all.x = TRUE,
                            by.x = "interval",
                            by.y = "interval")

merged_activity_ds[,steps := ifelse(is.na(steps),mean_of_steps,steps)]
```

## 6.3 Creating a new dataset that is equal to the original dataset but with the missing data filled in.
```{r}
activity_ds_filled <- merged_activity_ds[,c("steps", "date", "interval")]
```

## 6.4 Making a histogram on imputted data
```{r}
total_num_of_steps_filled <- activity_ds_filled[,.(sum_of_steps = sum(steps)),date]
```

# 7. Building histogram of the total number of steps taken each day after missing values are imputed
```{r}
hist(total_num_of_steps_filled$sum_of_steps, 
     main = "Distribution of total number of steps per day (NA removed)",
     xlab = "Total number of steps")

mean_steps_across_all_days_noNA <- mean(total_num_of_steps_filled$sum_of_steps,na.rm = TRUE)
mean_steps_across_all_days_noNA # 10766.19

median_steps_across_all_days_noNA <- median(total_num_of_steps_filled$sum_of_steps,na.rm = TRUE)
median_steps_across_all_days_noNA # 10766.19
```

As we see, median recalculated on inputted data is higher +1.19 steps comparing with original noninputted data. Mean stays the same.

# 8. Panel plot comparing the average number of steps taken per 5-minute interval across weekdays and weekends
## 8.1 Creating weekday/weekend var
```{r}
activity_ds_filled[,day_of_week := ifelse(
  weekdays(as.Date(date)) == "Saturday" | weekdays(as.Date(date)) == "Sunday", "weekend", "weekday")]

table(activity_ds_filled$day_of_week)
```

## 8.2 Creating panel plots
```{r}
mean_of_steps_weekday <- 
  activity_ds_filled[day_of_week == "weekday",.(mean_of_steps = mean(steps)), interval]

mean_of_steps_weekend <- 
  activity_ds_filled[day_of_week == "weekend",.(mean_of_steps = mean(steps)), interval]

par(mfrow=c(2,1))

plot(mean_of_steps_weekday$interval,
     mean_of_steps_weekday$mean_of_steps, type = "l",
     xlab = "interval", ylab = "steps on average")

plot(mean_of_steps_weekend$interval,
     mean_of_steps_weekend$mean_of_steps, type = "l",
     xlab = "interval", ylab = "steps on average")
```

As we could see, activity during weekends starts more later than in working days, and is a bit more intensive during the whole weekend days and evenings.
